---
title: "The Three Point Evolution of the Golden State Warriors" 
subtitle: "Coursework 1" 
author: "Abdullah Nasir" 
date: "19/03/2026" 
output: 
  html_document:
    toc: true
    toc_float: true
    theme: spacelab 
    highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
# This code will display the QMUL logo at the top right of the page
# Do not change this code
htmltools::img(src = knitr::image_uri("images/QMlogo.png"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0; padding:10px; width:20%;')
```

# Introduction

Modern basketball has undergone a dramatic transformation over the past
few decades. One of the clearest examples of this change is the rise of
the three-point shot, which is now a central part of offensive strategy
in the NBA and other Basketball leagues.

Few teams represent this evolution better than the Golden State
Warriors. Over the last decade, the team has helped to redefine modern
basketball through a style of play built around spacing, ball movement,
and long-range shooting. The success of players such as Stephen Curry
accelerated the league-wide shift towards three-point-heavy offences.

This project explores how the number of three-point attempts by the
Golden State Warriors has changed over time. Using NBA game data, I
analyse the evolution of the team’s three-point shooting using
time-series analysis. The data is then modelled using Meta’s Prophet
forecasting system to identify long-term trends and generate forecasts.

<img src="images/curry_3.avif" width="650"/>

# Dataset

The data used in this project comes from the **NBA Database (1947 –
Present)** available on **Kaggle**, which contains detailed game-level
statistics for NBA teams across many seasons.

Each row in the dataset represents the statistics for a team in a single
NBA game and includes variables such as field goals, rebounds, assists,
and three-point shooting. Because the dataset spans several decades, it
allows us to examine long-term changes in playing style.

For this analysis, I focus on the variable `threePointersAttempted`,
which records the number of three-point shots a team attempts in a game.
Since the goal of this analysis is to study how the Golden State
Warriors’ approach to three-point shooting has evolved over time, the
dataset is filtered to include only games played by the Warriors.

The dataset also includes the game date in the `gameDateTimeEst` column.
This variable allows the data to be organised as a time series so we can
analyse trends in three-point shooting across different eras of NBA
basketball.

```{r}
team_stats <- read.csv("data/TeamStatistics.csv")
```

# Preparing the Time Series

To analyse the Warriors’ three-point shooting over time, the dataset
must first be prepared as a time series.

After preparing the data as a time series, the dataset is filtered to
include only Golden State Warriors games. The game date variable is then
converted to a standard date format to serve as the time index for the
series.

In this project, we use the variable `threePointersAttempted` as the
time series value.

Because the NBA introduced the three-point line in 1979, the analysis is
restricted to games played since then. Additionally, data points with
missing values are removed to ensure the model can be estimated
correctly.

```{r}
warriors_stats <- subset(team_stats, teamName == "Warriors")

warriors_stats$gameDateTimeEst <- as.Date(warriors_stats$gameDateTimeEst)

warriors_ts <- data.frame(ds = warriors_stats$gameDateTimeEst, y = warriors_stats$threePointersAttempted)

warriors_ts <- subset(warriors_ts, ds >= as.Date("1979-01-01"))

warriors_ts <- na.omit(warriors_ts)
```

# Analysis

## Full Time Series Overview

We begin by visualising the full time series of Golden State Warriors
three-point attempts per game since the introduction of the three-point
line in 1979. This allows us to observe the overall long-term evolution
of three-point shooting within the team.

The three-point line was introduced to the NBA in the 1979–1980 season.
In its early years, teams attempted few shots from beyond the arc.
During the 1980s and early 1990s, most games featured fewer than ten
three-point attempts.

Over the decades, attempts steadily increased. This trend reflects a
broader strategic shift, with teams increasingly valuing the three-point
shot as an efficient scoring method.

Another notable aspect of the plot is the recurring gaps between groups
of observations. These represent the NBA off-season, when no
regular-season games occur. Each group thus indicates a single NBA
season.

The variation between games is also evident. Some games feature far more
three-point attempts due to factors such as opposing strategy, game
tempo, and player availability.

```{r}
plot(warriors_ts$ds, warriors_ts$y, type = "l", xlab = "Date", ylab = "Three-Point Attempts", main = "Golden State Warriors Three-Point Attempts Per Game (1979–Present)")
```

## The Modern Era: The Stephan Curry Effect

To better understand how the Warriors helped reshape modern basketball,
it is useful to zoom in on the more recent period beginning around 2014.

This period marks the rise of Stephen Curry and the Warriors’
championship teams, built around heavy three-point shooting and floor
spacing. As a result, during this era, the Warriors became one of the
league’s most influential teams, contributing to what is often called
the “three-point revolution.”

In the modern era, the spike in three-point attempts stands out. Many
games now feature thirty to forty attempts, which was extremely rare in
earlier decades.

```{r}
modern_era <- subset(warriors_ts, ds >= as.Date("2014-01-01"))

plot(
  modern_era$ds,
  modern_era$y,
  type = "l",
  xlab = "Date",
  ylab = "Three-Point Attempts",
  main = "Warriors Three-Point Attempts Per Game in the Modern Era"
)
```

The time series contains a few outliers. One, in particular, stands out:
the Warriors attempted 69 three-point shots in a single game.

This extreme value corresponds to a preseason game, which explains why
it appears slightly separated from the main clusters of regular-season
data in the time series. Preseason games are often used by teams to
experiment with tactics and evaluate offensive systems before the
competitive season begins.

For a team like the Warriors, whose system emphasises spacing and
perimeter shooting, this environment makes it easier to push those
strategies to the limit. While 69 attempts is clearly an extreme case,
it reflects the team's modern offensive philosophy and highlights how
dramatically three-point usage has evolved compared with earlier eras of
basketball.

## A Season in Focus

To better understand how shorter-term events can influence the time
series, it is useful to zoom in on a single NBA season. The 2019–2020
season provides an interesting case study because the Warriors
experienced major roster disruption and a long absence for Stephen Curry
early in the year.

```{r}
season_2019 <- subset(
  warriors_ts,
  ds >= as.Date("2019-10-01") & ds <= as.Date("2020-04-01")
)

plot(
  season_2019$ds,
  season_2019$y,
  type = "l",
  xlab = "Date",
  ylab = "Three-Point Attempts",
  main = "Warriors Three-Point Attempts During the 2019–2020 Season"
)
```

The plot highlights several noticeable changes in three-point attempts
throughout the season. At the start of the year, the Warriors attempted
around 35–45 three-point shots per game. This reflects the offensive
system built around perimeter shooting.

A period of decline appears towards the middle of October, following
Stephen Curry's injury, which broke his hand and ruled him out for most
of the season. Without their primary three-point shooter and offensive
leader, the Warriors’ offensive structure changed significantly, which
is reflected in the reduced number of three-point attempts over the
period.

In December, the team faced heavy roster instability, frequently relying
on younger or less experienced players. As a result, several games had
noticeably lower attempt totals, including the lowest values shown in
the plot.

In January, the series stabilises, as the Warriors return to their
typical 30–40 shots per game. During this period, players such as
D’Angelo Russell take on a larger offensive role, which increases
perimeter shot creation.

Finally, the spike at the end of the series occurred shortly before the
NBA suspended its 2019–2020 season in March 2020 due to the COVID-19
pandemic. Around this time, Stephen Curry briefly returned from injury,
and his return led to a short-term increase in offensive production and
three-point attempts.

Overall, this seasonal zoom shows how injuries, roster changes, and
external events can influence a sports time series. These factors can
affect team behaviour over short periods.

# Forecasting Model

To model the time series behaviour and generate future predictions, a
forecasting model is fitted to the data. For this analysis, the
**Prophet** model is used. Prophet is a forecasting tool designed for
time series data and is particularly useful for capturing long-term
trends and seasonal patterns.

The model is fitted to the Warriors' three-point attempts time series.

```{r setup}
library(prophet)
warriors_model <- prophet(warriors_ts)
```

After fitting the model, future dates are generated and predictions are
produced for those time points.

```{r}
future <- make_future_dataframe(warriors_model, periods = 365)

forecast <- predict(warriors_model, future)
```

The resulting forecast contains predicted values along with uncertainty
intervals for each future observation. These predictions can then be
visualised and interpreted in the following section.

# Results & Discussion

After fitting the Prophet model, the forecast can be visualised to
examine both the historical data and the model’s predicted future
values.

```{r}
plot(warriors_model, forecast)
```

The forecast plot displays historical observations and the model's
anticipated behaviour. Black points mark observed three-point attempts
per Warriors game, while the blue line represents the Prophet model's
estimated trend. The shaded area indicates the uncertainty interval.

The model captures the clear upward trend in three-point attempts over
time. In recent seasons, the Warriors have frequently attempted 40–50
three-point shots per game, reflecting the modern NBA’s strong emphasis
on perimeter shooting.

The forecast indicates this high volume of three-point attempts will
likely persist in the near future. However, sports data predictions
require careful interpretation. As shown earlier, factors such as
injuries, roster changes, and tactical shifts can significantly impact
team behaviour.

Moreover, much of the Warriors’ offensive strategy has been defined by
players such as Stephen Curry. If key players retire or teams adopt
different tactics, the pattern of three-point attempts could shift.
Therefore, the forecast should be considered an estimate based on
historical patterns rather than a firm prediction.

Overall, this analysis shows that time-series analysis help track shifts
in basketball strategy. Analysing the Golden State Warriors’ three-point
attempts highlights that perimeter shooting is increasingly vital to the
NBA, but forecasting future changes remains challenging.
